
\noindent 
This unit covers the following ideas. In preparation for the quiz and exam, make sure you have a review guide containing examples that explain and illustrate the following concepts.  
\begin{enumerate}
\item Explain how to setup and compute triple integrals, as well as how to interchange the bounds of integration. Use these ideas to find area and volume.
\item Explain how to change coordinate systems in integration, with an emphasis on cylindrical, and spherical coordinates. Explain what the Jacobian of a transformation is, and how to use it.
\item Use triple integrals to find physical quantities such as center of mass, radii of gyration, etc. for solid regions.
\item Explain how to use the Divergence theorem to compute the flux of a vector fields out of a closed surface.
\end{enumerate}

\newpage

%\stepcounter{unitday}
%\uday
%\normalsize

\section{Triple Integral Definition and Applications}

After completing this section you will:
\begin{itemize}
\item Be comfortable setting up triple integrals for a solid region (domain)
\item Be able to switch the order of integration for a triple integral
\end{itemize}


Consider the iterated integral $$\ds \int_{-3}^3 \int_0^{\sqrt{9-y^2}}\int_0^{9-x^2-y^2} 1dzdxdy.$$ This is an integral of the form $\iiint_D dV$, which means along some solid region $D$ in the plane, we are adding up little bits of volume. This integral will give the volume of a very familiar solid region in space. Not every triple integral will correspond to an easily drawn or visualized (even with a computer) space. However, this one does.
\begin{problem}
Before we compute this integral, let's visualize it. The parenthesis have been added to help you parse this first example.
$$\ds \int_{-3}^3 (\int_0^{\sqrt{9-y^2}} (\int_0^{9-x^2-y^2} 1dz) dx ) dy.$$
\begin{enumerate}
	\item State the bounds for each variable as inequalities.
	\item Sketch the region $D$ in space.  
	\item Compute the innermost integral, and compare the resulting/remaining double integral to the first exercise in the double integral unit. 
	\item Now evaluate the remaining integrals (though you might want to change coordinate systems before doing so).
\end{enumerate}
\end{problem}

When working with double integrals, there were two different ways to set up the bounds for our integrals, as $dA=dxdy=dydx$.  When working with triple integrals, there are six different ways to order the differentials, and therefore set up bounds for our integrals. $$dV=dxdydz = dxdzdy = dydxdz=dydzdx=dzdxdy=dzdydx$$ 


Consider again the iterated integral $$\ds \int_{-3}^3 \int_0^{\sqrt{9-y^2}}\int_0^{9-x^2-y^2} dzdxdy = \int_0^9\int_0^{\sqrt{9-z}}\int_{-\sqrt{9-x^2-z}}^{\sqrt{9-x^2-z}} dydxdz$$ These two integrals are equal, but required switching the order of the bounds. There are 4 other iterated integrals that are equal to these integral which can also be found by switching the order of the bounds. 

\begin{problem}
Set up the equivalent integrals using the differential orderings: $dydzdx$ and $dxdzdy$.  We'll look at the remaining 2 in class (though you're welcome to finish them and present them with your work). 
\end{problem}

\begin{problem}
Consider the iterated integral $$\int_{-1}^1\int_0^{1-x^2}\int_0^{y} dzdydx.$$
The bounds for this integral describe a region in space which satisfies the 3 inequalities $-1\leq x\leq 1$, $0\leq y\leq 1-x^2$, and $0\leq z\leq y$.
\begin{enumerate}
 \item Draw the solid domain $D$ in space described by the bounds of the iterated integral.
\end{enumerate} 
There are 5 other iterated integrals equivalent to this one. 
\begin{enumerate}[resume]
	\item Set up the integral that use the differential ordering: $dydxdz$ 
	\item Set up the integral that use the differential ordering: $dxdzdy$.  
\end{enumerate}
We'll create the other 3 in class (though you are welcome to include them as part of your presentation)
\end{problem}

%\newpage
%\Large
%\textbf{In-Class}\\
%\normalsize

\note{Need to include a review problem on sketching planes and finding their intercepts. Particularly with fractional equations}

\begin{problem}
 In each exercise below, you'll be given enough information to determine a solid domain $D$ in space. 
\begin{enumerate}[a)]
	\item Draw the solid $D$ and 
	\item then set up an iterated integral (pick any order you want) that would give the volume of $D$.  
\end{enumerate}
Note: You don't need to evaluate the integral, rather you just need to set them up.
\begin{enumerate}
 \item The region $D$ under the surface $z=y^2$, above the $xy$-plane, and bounded by the planes $y=-1$, $y=1$, $x=0$, and $x=4$.
 \item The region $D$ in the first octant that is bounded by the coordinate planes, the plane $y+z=2$, and the surface $x=4-y^2$.
 \item The pyramid $D$ in the first octant that is below the planes $\ds\frac{x}{3}+\frac{z}{2}=1$ and $\ds\frac{y}{5}+\frac{z}{2}=1$. [Hint: Don't let $z$ be the inside bound.]
 \item The region $D$ that is inside both right circular cylinders $x^2+z^2=1$ and $y^2+z^2=1$.
\end{enumerate}
\end{problem}

We can find average value, centroids, centers of mass, moments of inertia, and radii of gyration exactly as before. We just now need to integrate using three integrals, and replace $ds$, $dA$ or $d\sigma$, with $dV$.  
\begin{problem}
 Consider the triangular wedge $D$ that is in the first octant, bounded by the planes $\ds\frac{y}{7}+\frac{z}{5}=1$ and $x=12$. In the $yz$ plane, the wedge forms a triangle that passes through the points $(0,0,0)$,  $(0,7,0)$, and $(0,0,5)$.  [Hint: It may help to sketch a picture first]
\begin{enumerate}
	\item Set up integral formulas that would give the centroid $(\bar x,\bar y, \bar z)$ of $D$.  
	\item Actually compute the integrals for $\bar y$. 
	\item Now state $\bar x$ and $\bar z$ by using symmetry arguments.
\end{enumerate}
\end{problem}

\begin{problem}
 Consider the tetrahedron $D$ in the first octant that is underneath the plane that intersects the coordinate axes in the three point $(a,0,0)$, $(0,b,0)$ and $(0,0,c)$, where you can assume that $a,b,c>0$.
\begin{enumerate}[a)]
	\item An equation of an ellipse that passes through $(a,0)$ and $(0,b)$ is $\ds\frac{x^2}{a^2}+\frac{y^2}{b^2}=1$.  
	\item An equation of a line through these same two points is $\ds\frac{x}{a}+\frac{y}{b}=1$.  
	\item An equation of an ellipsoid through the three points $(a,0,0)$, $(0,b,0)$, and $(0,0,c)$ is $\ds\frac{x^2}{a^2}+\frac{y^2}{b^2}+\frac{z^2}{c^2}=1$.
\end{enumerate}
\begin{enumerate}
  \item  Guess an equation of the plane through these same three points, and then verify that your guess is correct by plugging the 3 points into your equation. This will provide you with an extremely fast way to get an equation of a plane.
  \item Set up an iterated integral that would give the volume of $D$.
  \item If the density is $\delta(x,y,z) = 3x+2yz$, set up iterated integrals that would give the mass $m$\bmw{ and moment of inertia $I_y$ about the $y$-axis}.
\end{enumerate}
\end{problem}

%
%\uday
%\normalsize


\section{Changing Coordinate Systems: The Jacobian}
After finishing this section, you should...
\begin{itemize}
\item Be able to change between standard coordinate systems for triple integrals:
\begin{itemize}
\item Spherical Coordinates
\item Cylindrical Coordinates
\end{itemize}
\end{itemize}

Just as we did with polar coordinates in two dimensions, we can compute a Jacobian for any change of coordinates in three dimensions.  We will focus on cylindrical and spherical coordinate systems. \\

Remember that the Jacobian of a transformation is found by first taking the derivative of the transformation, then finding the determinant, and finally computing the absolute value.\\

\begin{problem}
 The cylindrical change of coordinates is: 
\begin{align*}
	x&=r\cos\theta, y=r\sin\theta, z=z\\
	\text{or in vector form}&\  \\
	\vec C(r,\theta,z) &= (r\cos\theta,r\sin\theta, z)
\end{align*}  
The spherical change of coordinates is: 
\begin{align*}
	x&=\rho\sin\phi\cos\theta,\ y=\rho\sin\phi\sin\theta,\ z=\rho\cos\phi\\
	\text{or in vector form}&\ \\
	\vec S(\rho,\phi,\theta) &= (\rho\sin\phi\cos\theta,\rho\sin\phi\sin\theta,\rho\cos\phi). 
\end{align*}

\begin{enumerate}
 \item Verify that the Jacobian of the cylindrical transformation is $\ds\frac{\partial(x,y,z)}{\partial(r,\theta,z)} = |r|$.  
	\begin{itemize}
		\item If you want to make sure you don't have to use absolute values, what must you require?
	\end{itemize}
 \item Verify that the Jacobian of the spherical transformation is $\ds\frac{\partial(x,y,z)}{\partial(\rho,\phi,\theta)} = |\rho^2\sin\phi|$.  
	\begin{itemize}
		\item If you want to make sure you don't have to use absolute values, what must you require?
	\end{itemize}
\end{enumerate}
\end{problem}

The previous exercise shows us that, provided we require $r\geq0$ and $0\leq \phi\leq \pi$, we can write:
$$dV=dxdydz = rdrd\theta dz = \rho^2\sin\phi d\rho d\phi d\theta,$$
\vskip0.1in
\textbf{Cylindrical coordinates} are extremely useful for problems which involve:
\begin{itemize}\itemsep2pt
\item cylinders
\item paraboloids
\item cones
\end{itemize}

\textbf{Spherical coordinates} are extremely useful for problems which involve:
\begin{itemize}\itemsep2pt
\item cones
\item spheres
\end{itemize}

%\newpage
\vskip0.1in

\begin{problem}\marginpar{See \href{http://aleph.sagemath.org/?z=eJytjkEOgjAQRfecojvaMBKsYIxhOEpNFaNNgDZItO3pRaguSExcuJo_8_Jf5i57Gltw4GMW9eeL0h02sj3WkkzXPfElWsETJziRXU18hdkcBP-Q0OSLaqAV5lMjrCWut5FB1ZpGndRwMI0eNjUNFEcK1MKqgIIBde_gIYOcQXhxHqCNHA0Os3THIpN8d-Z_U_qXLnh_lfKFNb1d9YOyJ0VbdCM}{Sage}.}%
 The double cone $z^2=x^2+y^2$ has two halves.  Each half is called a nappe. Set up an integral in the coordinate system of your choice that would give the volume of the region that is between the $xy$ plane and the upper nappe of the double cone $z^2=x^2+y^2$, and between the cylinders $x^2+y^2=4$ and $x^2+y^2=16$.  Then evaluate the integral.
\end{problem}

\begin{problem}
 Set up an integral in the coordinate system of your choice that would give the volume of the solid ball that is inside the sphere $a^2=x^2+y^2+z^2$. Compute the integral to give a formula for the volume of a sphere of radius $a$.  \bmw{Then set up (don't evaluate) an iterated integral that would give the moment of inertia $I_x$ about the $x$-axis, if the density is a constant, so $\delta =c$.}
\end{problem}


\begin{problem}\marginpar{See \href{http://aleph.sagemath.org/?z=eJxty00OgyAQQOF9L-IQB6PSnxUnMWqImpTEBgqTFji900V35m3e5vuYAFXCjKUSF6_ty-92sTT73ZFaoWgd34EgTX2dp14gJJQKFU_-T8EWr4J1fcbvkq1ke0675oaPH27i033BRL8tNAdD1umhQ24UB9CiL84}{Sage}.}%
Find the volume of the solid domain $D$ in space which is above the cone $z=\sqrt{x^2+y^2}$ and below the paraboloid $z=6-x^2-y^2$. Use cylindrical coordinates to set up and then evaluate your integral.  \newline [Hint: You'll need to find where the surface intersect, as their intersection will help you determine the appropriate bounds.]
\end{problem}

For the next several exercises be sure to check that you've correctly swapped bounds by having Sage or WolframAlpha actually compute all of the integrals.

\begin{problem}
Consider the region $D$ in space that is inside both the sphere $x^2+y^2+z^2=9$ and the cylinder $x^2+y^2=4$.\newline [Hint: Start by drawing the region.]
\begin{enumerate}
 \item Set up an iterated integral in Cartesian (rectangular) coordinates that would give the volume of $D$. 
 \item Set up an iterated integral in cylindrical coordinates that would give the volume of $D$. 
\end{enumerate}
\end{problem}

\begin{problem}
Consider the region $D$ in space that is both inside the sphere $x^2+y^2+z^2=9$ and yet outside the cylinder $x^2+y^2=4$. \newline [Hint: Start by drawing the region.]
\begin{enumerate}
 \item Set up two iterated integrals in cylindrical coordinates that would give the volume of $D$. 
	\begin{enumerate}[a)]
		\item For the first integral use the order $dzdrd\theta$.  
		\item For the second, use the order $d\theta dr dz$.
	\end{enumerate}
 \item Set up an iterated integral in spherical coordinates that would give the volume of $D$. 
\end{enumerate}
\end{problem}



\begin{problem}
The integral $\ds\int_{0}^{\pi}\int_{0}^{1}\int_{\sqrt{3}r}^{\sqrt{4-r^2}}rdzdrd\theta$ represents the volume of solid domain $D$ in space. Set up integrals in both rectangular coordinates and spherical coordinates that would give the volume of the exact same region.
\end{problem}

\begin{problem}
The temperature at each point in space of a solid occupying the region {$D$}, which is the upper portion of the ball of radius 4 centered at the origin, is given by $T(x,y,z) = \sin(xy+z)$.  Set up an iterated integral formula that would give the average temperature.   
\end{problem}



%\uday
%\normalsize

\section{The Divergence Theorem}
\instructor{Old day -3- split here}
\instructor{There are several more commented out exercises in the .tex file that could be included, but are not well edited/tested yet.}
After completing this section you should be able to...
\begin{itemize}
\item Use the Divergence Theorem to compute flux across a surface
\end{itemize}

In the double integral chapter, we learned a way to greatly simplify flux computations when working with simple closed curves.  Green's theorem stated that $\int_C \vec F\cdot \vec n\ ds = \iint_R (M_x+N_y) dA$.  The divergence of $\vec F$ is the quantity $\text{div}(\vec F) = M_x+N_y$. We saw this in definition \ref{definition of flux density in 2D} on page \pageref{definition of flux density in 2D}, when we defined the divergence, or flux density, of a vector field $\vec F$ at a point $P$ to be the flux per unit area. We then stated that $\text{div}(\vec F)=M_x+N_y$. We more generally defined the divergence in the previous unit. This generalizes the simplification to higher dimensions, and is called the divergence theorem. We'll repeat the definition of divergence, with its theorem, for reference.

\begin{theorem}[Divergence Theorem]
Let $S$ be a closed surface whose interior is the solid domain $D$. Let $\vec n$ be an outward pointing unit normal vector to $S$. Suppose that $\vec F(x,y,z)$ is a continuously differentiable vector field on some open region that contains $D$. Then the outward flux of $\vec F$ across $S$ can be computed by adding up, along the entire solid $D$, the flux per unit volume (divergence).  Symbolically, the divergence theorem states
$$\iint_S\vec F\cdot \vec n d\sigma =  \iiint_D \vec \nabla \cdot \vec F dV = \iiint_D \left(M_x+N_y+P_z\right) dV $$
for $S$ a closed surface with interior $D$ and outward normal $\vec n$.
\end{theorem}

%In 3D, the flux of $\vec F$ across $S$, $\iint_S\vec F\cdot \vec n d\sigma$, is a measure of flow across $S$ where $\vec n$ is a continuous unit normal vector to $S$.  Flux density at $(x,y,z)$ is found by creating a sphere $S_a$ of radius $a$ centered at $(x,y,z)$ with interior volume $V_a$ and outward normal vector $\vec n$, and considering the quotient of flux per volume given by $\frac{1}{V_a}\iint_{S_a} \vec F \cdot \vec n d\sigma$. By computing $\ds \lim_{a\to 0}\frac{1}{V_a}\iint_{S_a} \vec F \cdot \vec n d\sigma$, we obtain the divergence of $\vec F$ at $(x,y,z)$, also called the flux density. In a future mathematics course, we could prove that the divergence equals
%\begin{align*}
%\text{div}\vec F(x,y,z) 
%&= \vec \nabla\cdot \vec F 
%= \left(\frac{\partial }{\partial x},\frac{\partial }{\partial y},\frac{\partial }{\partial z} \right)\cdot (M,N,P) \\
%&= \frac{\partial M}{\partial x}+\frac{\partial N}{\partial y}+\frac{\partial P}{\partial z} 
%= M_x+N_y+P_z 
%.
%\end{align*}

\begin{problem}
 Consider the exact same vector field and box as exercise \ref{boxflux}.  We then have the vector field $\vec F=(x+y,y,z) $  and $S$ is the surface of the cube in the first quadrant bounded by {$ x=2,y=3,z=5 $}.
\begin{enumerate}
 \item Compute the divergence of $\vec F$. I.E find $\text{div}(\vec F) = M_x+N_y+P_z$.
\end{enumerate}
The divergence theorem states that if $S$ is a closed surface (has an inside and an outside), and the inside of the surface is the solid domain $D$, then the flux of $\vec F$ outward across $S$ equals the triple integral
$$\iint_S\vec F\cdot \vec n\ d\sigma = \iint\int_D \text{div}(\vec F)dV.$$
\begin{enumerate}[resume]
	\item Use the divergence theorem to compute the flux of $\vec F$ across $S$.\newline [Hint: Just as the area is found by adding up little bits of area, which is what we mean by $A=\iint_R dA$, the volume is found by adding up little bits of volume.] 
\end{enumerate}
\end{problem}


%\begin{problem}
%In problem \ref{sphere surface area element}, we found 
%$$\vec n d\sigma = \vec r_\phi\times \vec r_\theta d\phi d\theta = a^2\sin \phi (\sin\phi\cos\theta,\sin\phi\sin\theta,\cos\phi)d\phi d\theta$$ for a sphere of radius $a$.  
%Use this to compute the outward flux of $$\ds \vec F=\frac{\left<-x,-y,-z\right>}{(x^2+y^2+z^2)^{3/2}} $$ across a sphere of radius $a$. You should get a negative number since the vector field has all arrows pointing in. [Hint: Remember that for a sphere of radius $a$ we have $a^2=x^2+y^2+z^2$. When you perform the dot product of $\vec F$ and $\vec n$, you'll save yourself a lot of time if you remember that $\vec u\cdot \vec u = |\vec u|^2$; the dot product of a vector with itself is the length squared.]
%
%\end{problem}
%
%
%\begin{problem}
%Repeat the previous problem, but this time don't use the formula from problem \ref{sphere surface area element}. In fact, you don't even need to parametrize the surface. Instead, if you are at the point $(x,y,z)$ on a sphere of radius $a$, give a formula for the outward pointing unit normal vector $\vec n$. Give this formula by only using a geometric argument.  Then find the outward flux of {$\ds \vec F=\frac{\left<-x,-y,-z\right>}{(x^2+y^2+z^2)^{3/2}} $} across a sphere of radius $a$. You should find that $\vec F\cdot \vec n$ simplifies to a constant, so that you never actually have to compute $d\sigma$. Then you can use known facts about the surface area of a sphere.
%
%\end{problem}


%In 3D, the flux of $\vec F$ across $S$, $\iint_S\vec F\cdot \vec n d\sigma$, is a measure of flow across $S$ where $\vec n$ is a continuous unit normal vector to $S$.  Flux density at $(x,y,z)$ is found by creating a sphere $S_a$ of radius $a$ centered at $(x,y,z)$ with interior volume $V_a$ and outward normal vector $\vec n$, and considering the quotient of flux per volume given by $\frac{1}{V_a}\iint_{S_a} \vec F \cdot \vec n d\sigma$. By computing $\ds \lim_{a\to 0}\frac{1}{V_a}\iint_{S_a} \vec F \cdot \vec n d\sigma$, we obtain the divergence of $\vec F$ at $(x,y,z)$, also called the flux density. In a future mathematics course, we could prove that the divergence equals
%\begin{align*}
%\text{div}\vec F(x,y,z) 
%&= \vec \nabla\cdot \vec F 
%= \left(\frac{\partial }{\partial x},\frac{\partial }{\partial y},\frac{\partial }{\partial z} \right)\cdot (M,N,P) \\
%&= \frac{\partial M}{\partial x}+\frac{\partial N}{\partial y}+\frac{\partial P}{\partial z} 
%= M_x+N_y+P_z 
%.
%\end{align*}

\note{I tried problems like this during the semester in the double integral section, and they didn't go very well.  Perhaps it was placement.  It was a waste of 20 minutes in the previous section.}

%\begin{problem}
%Compute the flux density at $(0,0,0)$ for the vector field $\vec F = \left<x,y,z\right>$. A sphere of radius $a$ has unit normal vector $\vec n = \frac{\left<x,y,z\right>}{|\left<x,y,z\right>|}$, so the flux is $\iint_S \left<x,y,z\right>\cdot \frac{\left<x,y,z\right>}{|\left<x,y,z\right>|}d\sigma 
%= \iint_S \sqrt{x^2+y^2+z^2}d\sigma 
%= a \iint_S d\sigma   = a4\pi a^2$, since the surface area of a sphere is $4\pi a^2$. The volume inside a sphere of radius $a$ is $\frac 43\pi a^3$. Hence $\lim_{a\to 0}\frac{3}{4\pi a^3}4\pi a^3 = 3$, which equals $\text{div}\vec F = M_x+N_y+P_z = 1+1+1=3$. 
%\end{problem}


%\begin{theorem}[Divergence Theorem]
%Let $S$ be a closed surface whose interior is the solid domain $D$. Let $\vec n$ be an outward pointing unit normal vector to $S$. Suppose that $\vec F(x,y,z)$ is a continuously differentiable vector field on some open region that contains $D$. Then the outward flux of $\vec F$ across $S$ can be computed by adding up, along the entire solid $D$, the flux per unit volume (divergence).  Symbolically, the divergence theorem states
%$$\iint_S\vec F\cdot \vec n d\sigma =  \iiint_D \vec \nabla \cdot \vec F dV = \iiint_D \left(M_x+N_y+P_z\right) dV $$
%for $S$ a closed surface with interior $D$ and outward normal $\vec n$.
%\end{theorem}


\begin{problem}
Let $S$ be the surface of the wedge in the first octant bounded by the planes $x=1$ and $\ds\frac{y}{2}+\frac{z}{3}=1$. Let $\vec F$ be the vector field $\left<x+3y^2,y^2-4x,2z+xy\right>$. Use the divergence theorem to compute the outward flux of $\vec F$ across $S$. Make sure you draw the wedge (you may find centroids and volume help complete this exercise rapidly).  
\end{problem}

\begin{problem}
Consider the vector field $\vec F = \left<yz,-xz,3xz\right>$.  Let $D$ be the solid region in space inside the cylinder of radius 4, above the plane $z=0$, and below the paraboloid $z=x^2+y^2$.  The surface $S$ consists of 3 portions, so computing the flux would require a rather time consuming process of parameterizing these 3 surfaces.  Instead, use the divergence theorem to compute the outward flux of $\vec F$ across the surface $S$.
\end{problem}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "215-problems"
%%% End: 